---
title: "Weather Events impact on People Health and Economics in the US"
author: "Carlos Correia"
date: "25 November 2014"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    theme: united
    toc: yes
    toc_depth: 4
---

## Synopsis
Weather Events have an impact on the Population Health as well on country Economics.  
The goal of this analysis is to identify which Weather Events are more harmful for population health and which Weather Events have the greatest economic consequences.  
For this analysis we use data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database with events from the year 1950 until end of November 2011.

## Data Processing

### Loading Libraries and defining global variables
```{r library}
library(dplyr)
library(reshape2)
library(ggplot2)
library(grid)
library(gridExtra)

## URL for the NOAA file data
fileURL      <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localZipFile <- "./data/storm_data.csv.bz2"
inputRmdFile <- "PA2.Rmd"
```

### Helper functions
#### downloadAndExtractZipFile(fileName)
The `downloadAndExtractZipFile` function will verify if the data file (`fileName`) is available, if not will download the file using _CURL_ method. 
```{r helperFunctions1}
## Download and Extract Zip data file
downloadAndExtractZipFile <- function(fileName){
  ## check if the data folder exists
  if(!file.exists("data")){
    dir.create("data")
  }
    
  ## check if zip file exists - if not download the zip file with the data
  if(!file.exists(fileName)){
    download.file(fileURL, destfile = fileName, method = "curl")
  }
}
```

#### readCSVFile(fileName, ...)
The `readCSVFile` function will check if the data file exists (`fileName`) and read the data using `read.csv` R function
```{r helperFunctions2}
## Reads the CSV data file
readCSVFile <- function(fileName, ...){
  if(! file.exists(fileName)){
    stop(paste("readDataFile: File ", fileName, " doesn't exist"))
  }
  
  print(paste("Reading file ", fileName))
  read.csv(fileName, ...)
}
```

#### cleanupEVTYPE(input)
The `cleanupEVTYPE` will receive as `input` the __EVTYPE__ from the NOAA data source and do some cleanup/fix typos
```{r helperFunctions3}
## Cleanup and normalize the EVTYPE
cleanupEVTYPE <- function(input){
  output <- input %>%
    toupper() %>%
    gsub(pattern = "^\\t+", replacement = "", perl = TRUE) %>%
    gsub(pattern = "^\\s+", replacement = "", perl = TRUE) %>%
    gsub(pattern = "^AVALANCE.+", replacement = "AVALANCHE", perl = TRUE) %>%
    gsub(pattern = "^BLIZZARD.+", replacement = "BLIZZARD", perl = TRUE) %>%
    gsub(pattern = "^COASTAL FLOOD.+", replacement = "COASTAL FLOOD", perl = TRUE) %>%
    gsub(pattern = "^COASTAL  FLOODING/EROSION$", replacement = "COASTAL FLOOD", perl = TRUE) %>%
    gsub(pattern = "^COASTALSTORM", replacement = "COASTAL STORM", perl = TRUE) %>%
    gsub(pattern = "^COLD.+", replacement = "COLD", perl = TRUE) %>%
    gsub(pattern = "^COOL AND WET$", replacement = "COLD", perl = TRUE) %>%
    gsub(pattern = "^DROUGHT.+", replacement = "DROUGHT", perl = TRUE) %>%
    gsub(pattern = "^DRY MIRCOBURST WINDS$", replacement = "DRY MIRCOBURST", perl = TRUE) %>%
    gsub(pattern = "^DUST DEVIL.+", replacement = "DUST DEVIL", perl = TRUE) %>%
    gsub(pattern = "^DUST STORM.+", replacement = "DUST STORM", perl = TRUE) %>%
    gsub(pattern = "^EXTREME COLD.+", replacement = "EXTREME COLD", perl = TRUE) %>%
    gsub(pattern = "^EXTREME WIND CHILL$", replacement = "EXTREME WINDCHILL", perl = TRUE) %>%
    gsub(pattern = "^FLASH FLOOD.+", replacement = "FLASH FLOOD", perl = TRUE) %>%
    gsub(pattern = "^FREEZE.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^FREEZING.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^GUSTY WIND", replacement = "GUSTY WINDS", perl = TRUE) %>%
    gsub(pattern = "^HEAT WAVE.+", replacement = "HEAT WAVE", perl = TRUE) %>%
    gsub(pattern = "^HEAVY RAIN.+", replacement = "HEAVY RAIN", perl = TRUE) %>%
    gsub(pattern = "^HEAVY SNOW.+", replacement = "HEAVY SNOW", perl = TRUE) %>%
    gsub(pattern = "^FLOOD.+", replacement = "FLOOD", perl = TRUE) %>%
    gsub(pattern = "^RIVER FLOOD.+", replacement = "FLOOD", perl = TRUE) %>%
    gsub(pattern = "^FREEZE.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^FROST.+", replacement = "FROST", perl = TRUE) %>%
    gsub(pattern = "^GUSTY WINDS.+", replacement = "GUSTY WINDS", perl = TRUE) %>%
    gsub(pattern = "^HAIL.+", replacement = "HAIL", perl = TRUE) %>%
    gsub(pattern = "^HEAVY SURF.+", replacement = "HEAVY SURF", perl = TRUE) %>%
    gsub(pattern = "^HIGH SURF.+", replacement = "HIGH SURF", perl = TRUE) %>%
    gsub(pattern = "^HIGH WIND.+", replacement = "HIGH WIND", perl = TRUE) %>%
    gsub(pattern = "^HIGH$", replacement = "HIGH WIND", perl = TRUE) %>% ## see REMARKS
    gsub(pattern = "^HURRICANE.+", replacement = "HURRICANE", perl = TRUE) %>%
    gsub(pattern = "^HYPOTHERMIA.+", replacement = "HYPOTHERMIA", perl = TRUE) %>%
    gsub(pattern = "^ICE.+", replacement = "ICE", perl = TRUE) %>%
    gsub(pattern = "^ICY ROADS.+", replacement = "ICE", perl = TRUE) %>%
    gsub(pattern = "^LANDSLIDE.+", replacement = "LANDSLIDE", perl = TRUE) %>%
    gsub(pattern = "^LIGHTNING.+", replacement = "LIGHTNING", perl = TRUE) %>%
    gsub(pattern = "^RECORD/EXCESSIVE HEAT$", replacement = "RECORD HEAT", perl = TRUE) %>%
    gsub(pattern = "^RIP CURRENT.+", replacement = "RIP CURRENT", perl = TRUE) %>%
    gsub(pattern = "^SNOW.+", replacement = "SNOW", perl = TRUE) %>%
    gsub(pattern = "^STORM SURGE.+", replacement = "STORM SURGE", perl = TRUE) %>%
    gsub(pattern = "^STRONG WIND.+", replacement = "STRONG WIND", perl = TRUE) %>%
    gsub(pattern = "^THUNDERSTORM.*", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^THUNDERTORM WINDS$", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^TSTM WIND.*", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^TORNADO.+", replacement = "TORNADO", perl = TRUE) %>%
    gsub(pattern = "^TROPICAL STORM.+", replacement = "TROPICAL STORM", perl = TRUE) %>%
    gsub(pattern = "^UNSEASONABLY WARM.+", replacement = "UNSEASONABLY WARM", perl = TRUE) %>%
    gsub(pattern = "^URBAN FLOOD.+", replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^URBAN SMALL.*", replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^URBAN/SMALL STREAM.*", replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^URBAN AND SMALL STREAM FLOODIN$", 
         replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^WATERSPOUT.+", replacement = "WATERSPOUT", perl = TRUE) %>%
    gsub(pattern = "^WILD.+", replacement = "WILDFIRE", perl = TRUE) %>%
    gsub(pattern = "^WIND.+", replacement = "WIND", perl = TRUE) %>%
    gsub(pattern = "^WINTER STORM.+", replacement = "WINTER STORM", perl = TRUE) %>%
    gsub(pattern = "^WINTER WEATHER.+", replacement = "WINTER WEATHER", perl = TRUE)
  
  output
}
```

#### calculateDMG(value, exp)
The `calculateDMG` function will calculate the monetary value of the damage using the `value` and `exp` as input. The `exp` can be:

- B or b: Billion
- M or m: Million
- K or k: Thousands
- H or h: Hundreds
- Any other option is ignored
```{r helperFunctions4}
## Calculates the DMG values by checking the EXP and update the DMG value
calculateDMG <- function(value, exp){
  output <- value
  
  if(as.numeric(value) > 0 && !is.null(exp)) {
    output <- switch(EXPR = as.character(exp),
           B =, b = value * 1000000000,
           M =, m = value * 1000000,
           K =, k = value * 1000,
           H =, h = value * 100,
           value)
  }

  output
}
```

### Loading Raw Data
Starting by calling the `downloadAndExtractZipFile` function to download the [NOAA data source][1] in case we don't have the file locally.  
Then we call the `readCSVFile` function to read the file and load into `data` data set variable
```{r dataLoad, cache = TRUE}
downloadAndExtractZipFile(localZipFile)

data <- readCSVFile(localZipFile)
str(data)
```

### Cleaning the Data
The raw data set has `r ncol(data)` columns, but for this analysis we only need the following columns: 

- __EVTYPE__ (_factor_): Weather Event Type
- __FATALITIES__ (_num_): Number of Fatalities
- __INJURIES__ (_num_): Number of Injuries
- __PROPDMG__ (_num_): Value for the property damage
- __PROPDMGEXP__ (_factor_): Exponential  for property damage (Billions, Millions, Thousands, Hundreds)
- __CROPDMG__ (_num_): Value for the crop damage
- __CROPDMGEXP__ (_factor_): Exponential  for crop damage (Billions, Millions, Thousands, Hundreds)

```{r tidyDataSummary, cache=TRUE}
tidyData <- data[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
str(tidyData)
summary(tidyData)
```

#### TidyData for Weather Events impacts on Population Health
Using the `tidyData` data set the the input, we ignore the rows where we don't have Fatalities and Injuries since we want to get the Weather Events with most Health impact. After we select only the 3 columns we need - `EVTYPE`, `FATALITIES` and `INJURIES`, do some cleanup to the `EVTYPE` factor calling the `cleanupEVTYPE` function, group the data by `EVTYPE` and then calculate the sum of Fatalities and Injures for each Weather Event (`EVTYPE`). Since we want the most impact, we sort the total Fatalities and Injures descendant, pick the top 10 and apply the `melt` R function to prepare the data for the plot.
```{r tidyDataHarmful}
tidyDataHarmful <- tidyData[!(tidyData$FATALITIES == 0 & tidyData$INJURIES == 0), ] %>%
  select(EVTYPE, FATALITIES, INJURIES) %>%
  mutate(EVTYPE = cleanupEVTYPE(EVTYPE)) %>%
  group_by(EVTYPE) %>%
  summarise(Fatality = sum(FATALITIES), Injury = sum(INJURIES)) %>%
  arrange(desc(Fatality), desc(Injury)) %>%
  head(n = 10) %>%
  melt(id = c("EVTYPE"), variable.name = "Type")

str(tidyDataHarmful)
```

```{r summaryTidyDataHarmful}
summary(tidyDataHarmful)
```

#### TidyData for Weather Events impacts on Economics
From the `tidyData`data set, we ignore the rows where there isn't Property nor Crop damage value because we want to get the Weather Events with most Economical impact. After we select only the 5 columns we need - `EVTYPE`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG` and `CROPDMGEXP`, do some cleanup to the `EVTYPE` factor calling the `cleanupEVTYPE` function. In other to get the real monetary value for Property or Crop damage, we call the function `calculateDMG`. Then we group the data by `EVTYPE` and then calculate the total of Properties and Crop monetary damage for each Weather Event (`EVTYPE`).
```{r tidyDataDMG}
tidyDataDMG <- tidyData[!(tidyData$PROPDMG == 0 & tidyData$CROPDMG == 0), ] %>%
  select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  mutate(EVTYPE = cleanupEVTYPE(EVTYPE)) %>%
  mutate(FINALPROPDMG = mapply(calculateDMG, PROPDMG, PROPDMGEXP), 
         FINALCROPDMG = mapply(calculateDMG, CROPDMG, CROPDMGEXP)) %>%
  group_by(EVTYPE) %>%
  summarise(Property = sum(FINALPROPDMG), 
            Crop = sum(FINALCROPDMG))

str(tidyDataDMG)
```

```{r summaryTidyDataDMG}
summary(tidyDataDMG)
```

## Results
### Weather Event Impact on US Population Health
Data from the data set of the most impact Weather Event of Population Health
```{r tidyDataHarmfulResult}
tidyDataHarmful
```

Now we plot to show the which Weather Event has the bigger impact on the US Population Health
```{r plotHarmfull, fig.width=9, fig.cap="Fig. 1) Top 10 Weather Events on Population Health"}
plotHarmfull <- 
    ggplot(tidyDataHarmful, aes(x = reorder(EVTYPE, -value), y = value/1000, fill = Type)) +
    geom_bar(stat="identity", position="dodge") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Number of people (thousands)", 
        x = "",
      title = "Impact of Weather Events on the US Population Health")
    
print(plotHarmfull)
```

### Weather Events Impact on Economic
The first 5 rows of the tidy data set for the Weather Events Impact on Economic analysis
```{r tidyDataDMGResult}
head(tidyDataDMG, n=5)
```

Since we want the most Economical impact, first we will do that for Properties by creating a new data set `plotPropDMGData` using the `tidyDataDMG` data set as the start, pick only the `EVTYPE` and `Property` column, sort descendant and get the top 10. 
```{r tidyDataDMGProp}
plotPropDMGData <- tidyDataDMG %>%
  select(EVTYPE, Property) %>%
  arrange(desc(Property)) %>%
  head(n=10)

plotPropDMGData
```

Now we do exactly the same we did in the previous steps but now for CROP
```{r tidyDataDMGCrop}
plotCropDMGData <- tidyDataDMG %>%
  select(EVTYPE, Crop) %>%
  arrange(desc(Crop)) %>%
  head(n=10)

plotCropDMGData
```

And we plot to show the Weather Impact on the US Economic with Property and Crop damage together
```{r plotDMGPROPAndCROP, fig.width=9, fig.cap="Fig. 2) Top 10 Weather Events Economical Impact on (left) Property (right) Crop"}
plotDMGPROP <- 
    ggplot(plotPropDMGData, aes(x = reorder(EVTYPE, -Property), 
                                y = Property/10^9)) +
    geom_bar(stat="identity", fill="blue") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Billions of Dollars", 
        x = "",
      title = "Property")

plotCROPDMG <- 
    ggplot(plotCropDMGData, aes(x = reorder(EVTYPE, -Crop), 
                                y = Crop/10^9)) +
    geom_bar(stat="identity", fill="blue") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Billions of Dollars", 
        x = "",
      title = "Crop")

grid.arrange(plotDMGPROP, plotCROPDMG, ncol = 2,
             main = "Economic impact of Weather Events in the US")

```
  
  
Now we prepare a new data set where we will pick the top 20 Weather Events with the highest Property and Crop monetary damage 
```{r tidyDataDMGPropCrop}
plotTotalDMGData <- tidyDataDMG %>%
  arrange(desc(Property), desc(Crop)) %>%
  head(n=20) %>%
  melt(id=c("EVTYPE"), variable.name = "Type")

plotTotalDMGData
```

We plot to show which Weather Event has the most Impact on country Economic (Property and Crop)
```{r plotTotalDMG, fig.width=9, fig.cap="Fig. 3) Top 20 Weather Events Economical Impact (Property and Crop)"}

plotTOTALDMG <- 
    ggplot(plotTotalDMGData, aes(x = reorder(EVTYPE, -value), 
                                 y = value/1000000000, fill = Type)) +
    geom_bar(stat="identity") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Billions of Dollars", 
        x = "",
      title = "Economic impact of Weather Events in the US")

print(plotTOTALDMG)
```

## Conclusions
With this analysis we conclude that:

* `r tidyDataHarmful[1, "EVTYPE"]` is the Weather Event with the biggest impact on the US Population Health
* `r plotPropDMGData[1, "EVTYPE"]` is the Weather Event with biggest Economical impact on the Properties
* `r plotCropDMGData[1, "EVTYPE"]` is the Weather Event with biggest Economical impact on the Crop
* `r plotTotalDMGData[1, "EVTYPE"]` and `r plotTotalDMGData[2, "EVTYPE"]` are the Weather Events with biggest Economical impact in the US (Property and Crop)

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "NOAA Data Source"
