---
title: "RepData - Peer Assessment 2"
author: "Carlos Correia"
date: "22 November 2014"
output:
  html_document:
    keep_md: yes
---

##Synopsis

##Data Processing

### Loading Libraries and defining global variables
```{r library}
library(dplyr)
library(reshape2)
library(ggplot2)

fileURL      <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localZipFile <- "./data/storm_data.csv.bz2"
inputRmdFile <- "PA2.Rmd"
```

### Helper functions
```{r helperFunctions}
## Download and Extract Zip data file
downloadAndExtractZipFile <- function(fileName){
  ## check if the data folder exists
  if(!file.exists("data")){
    dir.create("data")
  }
    
  ## check if zip file exists - if not download the zip file with the data
  if(!file.exists(fileName)){
    download.file(fileURL, destfile = fileName, method = "curl")
  }
}

## Reads the CSV data file
readCSVFile <- function(fileName, ...){
  if(! file.exists(fileName)){
    stop(paste("readDataFile: File ", fileName, " doesn't exist"))
  }
  
  print(paste("Reading file ", fileName))
  read.csv(fileName, ...)
}

cleanupEVTYPE <- function(input){
  output <- input %>%
    toupper() %>%
    gsub(pattern = "^AVALANCE", replacement = "AVALANCHE", perl = TRUE) %>%
    gsub(pattern = "^COASTAL FLOOD.+", replacement = "COASTAL FLOOD", perl = TRUE) %>%
    gsub(pattern = "^COASTALSTORM", replacement = "COASTAL STORM", perl = TRUE) %>%
    gsub(pattern = "^COLD.+", replacement = "COLD", perl = TRUE) %>%
    gsub(pattern = "^DROUGHT.+", replacement = "DROUGHT", perl = TRUE) %>%
    gsub(pattern = "^DRY MIRCOBURST WINDS$", replacement = "DRY MIRCOBURST", perl = TRUE) %>%
    gsub(pattern = "^EXTREME COLD.+", replacement = "EXTREME COLD", perl = TRUE) %>%
    gsub(pattern = "^FLASH FLOOD.+", replacement = "FLASH FLOOD", perl = TRUE) %>%
    gsub(pattern = "^FREEZE.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^GUSTY WIND", replacement = "GUSTY WINDS", perl = TRUE) %>%
    gsub(pattern = "^HEAT WAVE.+", replacement = "HEAT WAVE", perl = TRUE) %>%
    gsub(pattern = "^HEAVY RAIN.+", replacement = "HEAVY RAIN", perl = TRUE) %>%
    gsub(pattern = "^HEAVY SNOW.+", replacement = "HEAVY SNOW", perl = TRUE) %>%
    gsub(pattern = "^FLOOD.+", replacement = "FLOOD", perl = TRUE) %>%
    gsub(pattern = "^RIVER FLOOD.+", replacement = "FLOOD", perl = TRUE) %>%
    gsub(pattern = "^FREEZE.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^GUSTY WINDS.+", replacement = "GUSTY WINDS", perl = TRUE) %>%
    gsub(pattern = "^HEAVY SURF.+", replacement = "HEAVY SURF", perl = TRUE) %>%
    gsub(pattern = "^HIGH WIND.+", replacement = "HIGH WIND", perl = TRUE) %>%
    gsub(pattern = "^HIGH$", replacement = "HIGH WIND", perl = TRUE) %>% ## see REMARKS
    gsub(pattern = "^HURRICANE.+", replacement = "HURRICANE", perl = TRUE) %>%
    gsub(pattern = "^HYPOTHERMIA.+", replacement = "HYPOTHERMIA", perl = TRUE) %>%
    gsub(pattern = "^ICE.+", replacement = "ICE", perl = TRUE) %>%
    gsub(pattern = "^ICY ROADS.+", replacement = "ICE", perl = TRUE) %>%
    gsub(pattern = "^LANDSLIDE.+", replacement = "LANDSLIDE", perl = TRUE) %>%
    gsub(pattern = "^LIGHTNING.+", replacement = "LIGHTNING", perl = TRUE) %>%
    gsub(pattern = "^RECORD/EXCESSIVE HEAT$", replacement = "RECORD HEAT", perl = TRUE) %>%
    gsub(pattern = "^RIP CURRENT.+", replacement = "RIP CURRENT", perl = TRUE) %>%
    gsub(pattern = "^SNOW.+", replacement = "SNOW", perl = TRUE) %>%
    gsub(pattern = "^STORM SURGE.+", replacement = "STORM SURGE", perl = TRUE) %>%
    gsub(pattern = "^STRONG WIND.+", replacement = "STRONG WIND", perl = TRUE) %>%
    gsub(pattern = "^THUNDERSTORM.*", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^THUNDERTORM WINDS$", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^TSTM WIND.*", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^TORNADO.+", replacement = "TORNADO", perl = TRUE) %>%
    gsub(pattern = "^TROPICAL STORM.+", replacement = "TROPICAL STORM", perl = TRUE) %>%
    gsub(pattern = "^UNSEASONABLY WARM.+", replacement = "UNSEASONABLY WARM", perl = TRUE) %>%
    gsub(pattern = "^URBAN AND SMALL STREAM FLOODIN$", 
         replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^WATERSPOUT.+", replacement = "WATERSPOUT", perl = TRUE) %>%
    gsub(pattern = "^WILD.+", replacement = "WILDFIRE", perl = TRUE) %>%
    gsub(pattern = "^WIND.+", replacement = "WIND", perl = TRUE) %>%
    gsub(pattern = "^WINTER STORM.+", replacement = "WINTER STORM", perl = TRUE) %>%
    gsub(pattern = "^WINTER WEATHER.+", replacement = "WINTER WEATHER", perl = TRUE)
  
  output
}

calculateDMG <- function(value, exp){
  output <- value
  if(as.numeric(value) > 0) {
    output <- switch(EXPR = exp,
           B =, b = value * 1000000000,
           M =, m = value * 1000000,
           K =, k = value * 1000,
           H =, h = value * 100,
           -1)
  }
  
  output
}
```

### Load Raw Data
```{r dataLoad, cache = TRUE}
downloadAndExtractZipFile(localZipFile)

data <- readCSVFile(localZipFile)

## summary(data)
```

### Tidy Data
```{r tidyDataHarmfull, cache=FALSE}

tidyDataHarmful <- data[!(data$FATALITIES == 0 & data$INJURIES == 0), ] %>%
  select(EVTYPE, FATALITIES, INJURIES) %>%
  mutate(EVTYPE = cleanupEVTYPE(EVTYPE)) %>%
  group_by(EVTYPE) %>%
  summarise(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES)) %>%
  arrange(desc(FATALITIES), desc(INJURIES)) %>%
  head(n = 10) %>%
  melt(id=c("EVTYPE"), variable.name = "TYPE")

str(tidyDataHarmful)
```

```{r summaryTidyDataHarmful}
summary(tidyDataHarmful)
```

```{r tidyDataDMG, cache=FALSE}

tidyDataDMG <- data[!(data$PROPDMG == 0 & data$CROPDMG == 0), ] %>%
  select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS) %>%
  mutate(EVTYPE = cleanupEVTYPE(EVTYPE)) %>%
  mutate(FINALPROPDMG = calculateDMG(PROPDMG, PROPDMGEXP), 
         FINALCROPDMG = calculateDMG(CROPDMG, CROPDMGEXP))
  


str(tidyDataDMG)
```

```{r summaryTidyDataDMG}
summary(tidyDataDMG)
```

##Results


####Questions
1) Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r plotHarmfull}

plotHarmfull <- 
    ggplot(tidyDataHarmful, aes(x = reorder(EVTYPE, -value), y = value, fill = TYPE)) +
    geom_bar(stat="identity", position="dodge") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Number of people", 
        x = "Type of Events",
      title = "TOP 10 of events more harmfull")
    
print(plotHarmfull)
```


2) Across the United States, which types of events have the greatest economic consequences?



### Software Details used on this project
```{r softwareDetails}
sessionInfo()
```

