---
title: "Weather Events impact on People Health and Economics in the US"
author: "Carlos Correia"
date: "25 November 2014"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    theme: united
    toc: yes
    toc_depth: 4
---

## Synopsis
Weather Events have an impact on the Population Health as well on country Economics.  
The goal of this analysis is to identify which Weather Events are more harmful for population health and which Weather Events have the greatest economic consequences.  
For this analysis we use data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database with events from the year 1950 until end of November 2011.

## Data Processing

### Loading Libraries and defining global variables
```{r library}
library(dplyr)
library(reshape2)
library(ggplot2)
library(grid)
library(gridExtra)

## URL for the NOAA file data
fileURL      <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localZipFile <- "./data/storm_data.csv.bz2"
inputRmdFile <- "PA2.Rmd"
```

### Loading Raw Data

The data was downloaded from the [NOAA data source][1] using the _CURL_ method into a folder called _data_. The download is only done when the data file is not already present in the working directory.

```{r helperFunctions1}
## Download BZ2 CSV data file
downloadAndExtractZipFile <- function(fileName){
  ## check if the data folder exists
  if(!file.exists("data")){
    dir.create("data")
  }
    
  ## check if zip file exists - if not download the zip file with the data
  if(!file.exists(fileName)){
    download.file(fileURL, destfile = fileName, method = "curl")
  }
}
```

The downloaded compressed _CSV_ file was then loaded directly into a dataset - `data`.

```{r helperFunctions2}
## Reads the CSV data file
readCSVFile <- function(fileName, ...){
  if(! file.exists(fileName)){
    stop(paste("readDataFile: File ", fileName, " doesn't exist"))
  }
  
  print(paste("Reading file ", fileName))
  read.csv(fileName, ...)
}
```

The resulting dataset is summarised below.

```{r dataLoad, cache = TRUE}
downloadAndExtractZipFile(localZipFile)

data <- readCSVFile(localZipFile)
str(data)
```

### Cleaning the Data
The raw data set has `r ncol(data)` columns, but for this analysis we only need the following columns: 

- __EVTYPE__ (_factor_): Weather Event Type
- __FATALITIES__ (_num_): Number of Fatalities
- __INJURIES__ (_num_): Number of Injuries
- __PROPDMG__ (_num_): Value for the property damage
- __PROPDMGEXP__ (_factor_): Exponential  for property damage (Billions, Millions, Thousands, Hundreds)
- __CROPDMG__ (_num_): Value for the crop damage
- __CROPDMGEXP__ (_factor_): Exponential  for crop damage (Billions, Millions, Thousands, Hundreds)

```{r tidyDataSummary, cache=TRUE}
tidyData <- data[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
str(tidyData)
summary(tidyData)
```

The `EVTYPE` column needed further cleanup as some values in this cleanup are either redundant, confusing or miswritten. 
The cleanup was done based on regular expression patterns and it attempted to correct typos and group event types based on similarities. This cleanup was not exhaustive but it proved to be enough for the purposes of this analysis.

```{r helperFunctions3}
## Cleanup and normalize the EVTYPE
cleanupEVTYPE <- function(input){
  output <- input %>%
    toupper() %>%
    gsub(pattern = "^\\t+", replacement = "", perl = TRUE) %>%
    gsub(pattern = "^\\s+", replacement = "", perl = TRUE) %>%
    gsub(pattern = "^AVALANCE.+", replacement = "AVALANCHE", perl = TRUE) %>%
    gsub(pattern = "^BLIZZARD.+", replacement = "BLIZZARD", perl = TRUE) %>%
    gsub(pattern = "^COASTAL FLOOD.+", replacement = "COASTAL FLOOD", perl = TRUE) %>%
    gsub(pattern = "^COASTAL  FLOODING/EROSION$", replacement = "COASTAL FLOOD", perl = TRUE) %>%
    gsub(pattern = "^COASTALSTORM", replacement = "COASTAL STORM", perl = TRUE) %>%
    gsub(pattern = "^COLD.+", replacement = "COLD", perl = TRUE) %>%
    gsub(pattern = "^COOL AND WET$", replacement = "COLD", perl = TRUE) %>%
    gsub(pattern = "^DROUGHT.+", replacement = "DROUGHT", perl = TRUE) %>%
    gsub(pattern = "^DRY MIRCOBURST WINDS$", replacement = "DRY MIRCOBURST", perl = TRUE) %>%
    gsub(pattern = "^DUST DEVIL.+", replacement = "DUST DEVIL", perl = TRUE) %>%
    gsub(pattern = "^DUST STORM.+", replacement = "DUST STORM", perl = TRUE) %>%
    gsub(pattern = "^EXTREME COLD.+", replacement = "EXTREME COLD", perl = TRUE) %>%
    gsub(pattern = "^EXTREME WIND CHILL$", replacement = "EXTREME WINDCHILL", perl = TRUE) %>%
    gsub(pattern = "^FLASH FLOOD.+", replacement = "FLASH FLOOD", perl = TRUE) %>%
    gsub(pattern = "^FREEZE.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^FREEZING.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^GUSTY WIND", replacement = "GUSTY WINDS", perl = TRUE) %>%
    gsub(pattern = "^HEAT WAVE.+", replacement = "HEAT WAVE", perl = TRUE) %>%
    gsub(pattern = "^HEAVY RAIN.+", replacement = "HEAVY RAIN", perl = TRUE) %>%
    gsub(pattern = "^HEAVY SNOW.+", replacement = "HEAVY SNOW", perl = TRUE) %>%
    gsub(pattern = "^FLOOD.+", replacement = "FLOOD", perl = TRUE) %>%
    gsub(pattern = "^RIVER FLOOD.+", replacement = "FLOOD", perl = TRUE) %>%
    gsub(pattern = "^FREEZE.+", replacement = "FREEZE", perl = TRUE) %>%
    gsub(pattern = "^FROST.+", replacement = "FROST", perl = TRUE) %>%
    gsub(pattern = "^GUSTY WINDS.+", replacement = "GUSTY WINDS", perl = TRUE) %>%
    gsub(pattern = "^HAIL.+", replacement = "HAIL", perl = TRUE) %>%
    gsub(pattern = "^HEAVY SURF.+", replacement = "HEAVY SURF", perl = TRUE) %>%
    gsub(pattern = "^HIGH SURF.+", replacement = "HIGH SURF", perl = TRUE) %>%
    gsub(pattern = "^HIGH WIND.+", replacement = "HIGH WIND", perl = TRUE) %>%
    gsub(pattern = "^HIGH$", replacement = "HIGH WIND", perl = TRUE) %>% ## see REMARKS
    gsub(pattern = "^HURRICANE.+", replacement = "HURRICANE", perl = TRUE) %>%
    gsub(pattern = "^HYPOTHERMIA.+", replacement = "HYPOTHERMIA", perl = TRUE) %>%
    gsub(pattern = "^ICE.+", replacement = "ICE", perl = TRUE) %>%
    gsub(pattern = "^ICY ROADS.+", replacement = "ICE", perl = TRUE) %>%
    gsub(pattern = "^LANDSLIDE.+", replacement = "LANDSLIDE", perl = TRUE) %>%
    gsub(pattern = "^LIGHTNING.+", replacement = "LIGHTNING", perl = TRUE) %>%
    gsub(pattern = "^RECORD/EXCESSIVE HEAT$", replacement = "RECORD HEAT", perl = TRUE) %>%
    gsub(pattern = "^RIP CURRENT.+", replacement = "RIP CURRENT", perl = TRUE) %>%
    gsub(pattern = "^SNOW.+", replacement = "SNOW", perl = TRUE) %>%
    gsub(pattern = "^STORM SURGE.+", replacement = "STORM SURGE", perl = TRUE) %>%
    gsub(pattern = "^STRONG WIND.+", replacement = "STRONG WIND", perl = TRUE) %>%
    gsub(pattern = "^THUNDERSTORM.*", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^THUNDERTORM WINDS$", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^TSTM WIND.*", replacement = "THUNDERSTORM WIND", perl = TRUE) %>%
    gsub(pattern = "^TORNADO.+", replacement = "TORNADO", perl = TRUE) %>%
    gsub(pattern = "^TROPICAL STORM.+", replacement = "TROPICAL STORM", perl = TRUE) %>%
    gsub(pattern = "^UNSEASONABLY WARM.+", replacement = "UNSEASONABLY WARM", perl = TRUE) %>%
    gsub(pattern = "^URBAN FLOOD.+", replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^URBAN SMALL.*", replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^URBAN/SMALL STREAM.*", replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^URBAN AND SMALL STREAM FLOODIN$", 
         replacement = "URBAN/SML STREAM FLD", perl = TRUE) %>%
    gsub(pattern = "^WATERSPOUT.+", replacement = "WATERSPOUT", perl = TRUE) %>%
    gsub(pattern = "^WILD.+", replacement = "WILDFIRE", perl = TRUE) %>%
    gsub(pattern = "^WIND.+", replacement = "WIND", perl = TRUE) %>%
    gsub(pattern = "^WINTER STORM.+", replacement = "WINTER STORM", perl = TRUE) %>%
    gsub(pattern = "^WINTER WEATHER.+", replacement = "WINTER WEATHER", perl = TRUE)
  
  output
}
```

#### TidyData for analysing the impact of weather events on Population Health

To create the tidy dataset for the analysis of impact on population health rows that don't have fatalities or injuries were ignored. Similarly, only columns related to population health were included - `EVTYPE`, `FATALITIES` and `INJURIES`. The actual cleanup function for `EVTYPE` was also applied at this stage.

The data was then grouped by `EVTYPE` and the sum of fatalities and injuries was calculated for each event type. Since we are looking at the events with the most impact, the total sum was sorted in descending order. The top 10 was then picked from the descending dataset and the `melt` function was applied to prepare the data for ploting.

```{r tidyDataHarmful}
tidyDataHarmful <- tidyData[!(tidyData$FATALITIES == 0 & tidyData$INJURIES == 0), ] %>%
  select(EVTYPE, FATALITIES, INJURIES) %>%
  mutate(EVTYPE = cleanupEVTYPE(EVTYPE)) %>%
  group_by(EVTYPE) %>%
  summarise(Fatality = sum(FATALITIES), Injury = sum(INJURIES)) %>%
  arrange(desc(Fatality), desc(Injury)) %>%
  head(n = 10) %>%
  melt(id = c("EVTYPE"), variable.name = "Type")

str(tidyDataHarmful)
```

```{r summaryTidyDataHarmful}
summary(tidyDataHarmful)
```

#### TidyData for analysing the impact of weather events on Economics
To create the tidy dataset for the analysis of impact on the economy rows that don't have crop or property damage were ignored. Similarly, only columns related to economic impact were included - `EVTYPE`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG` and `CROPDMGEXP`. The actual cleanup function for `EVTYPE` was also applied at this stage.

In other to get the real monetary value for property or crop damage, the actual damage value - `PROPDMG` and `CROPDMG` - needs to multiplied by the exponent set in the corresponding exponent columns - `PROPDMGEXP` and `CROPDMGEXP`.

For the purpose of this study we assume the exponents to have the following values only. Any other option is ignored.

- B or b: Billion
- M or m: Million
- K or k: Thousands
- H or h: Hundreds

```{r helperFunctions4}
## Calculates the DMG values by checking the EXP and update the DMG value
calculateDMG <- function(value, exp){
  output <- value
  
  if(as.numeric(value) > 0 && !is.null(exp)) {
    output <- switch(EXPR = as.character(exp),
           B =, b = value * 1000000000,
           M =, m = value * 1000000,
           K =, k = value * 1000,
           H =, h = value * 100,
           value)
  }

  output
}
```

The data was then grouped by `EVTYPE` and the sum of property and crop damages was calculated for each event type.

Since we are looking at the events with the most impact, the total sum was sorted in descending order and the top 10 was then picked.

```{r tidyDataDMG}
tidyDataDMG <- tidyData[!(tidyData$PROPDMG == 0 & tidyData$CROPDMG == 0), ] %>%
  select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  mutate(EVTYPE = cleanupEVTYPE(EVTYPE)) %>%
  mutate(FINALPROPDMG = mapply(calculateDMG, PROPDMG, PROPDMGEXP), 
         FINALCROPDMG = mapply(calculateDMG, CROPDMG, CROPDMGEXP)) %>%
  group_by(EVTYPE) %>%
  summarise(Property = sum(FINALPROPDMG), 
            Crop = sum(FINALCROPDMG))

str(tidyDataDMG)
```

```{r summaryTidyDataDMG}
summary(tidyDataDMG)
```

## Results

### Impact of Weather Events on US Population Health

The following shows the dataset that contains the most harmful weather events in terms of population health. The dataset is small enough in this case to be able to easily display all of it.

```{r tidyDataHarmfulResult}
tidyDataHarmful
```

Plotting the dataset above gives a better overview of which events have the most impact on US population health.

```{r plotHarmfull, fig.width=9, fig.cap="Fig. 1) Top 10 Weather Events with impact on Population Health"}
plotHarmfull <- 
    ggplot(tidyDataHarmful, aes(x = reorder(EVTYPE, -value), y = value/1000, fill = Type)) +
    geom_bar(stat="identity", position="dodge") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Number of people (thousands)", 
        x = "",
      title = "Impact of Weather Events on the US Population Health")
    
print(plotHarmfull)
```

### Economic Impact of Weather Events

The following shows the first 5 rows of the dataset that contains the most harmful weather events in terms of economic cost. In this case the dataset is quite big so only the first 5 rows were shown for illustration purposes.

```{r tidyDataDMGResult}
head(tidyDataDMG, n=5)
```

To show the impact on property and crop damage separately the original tidy dataset with the economical impact was divided in 2 new datasets:

- A dataset to analyse property damage with only the `EVTYPE` and `Property` columns

```{r tidyDataDMGProp}
plotPropDMGData <- tidyDataDMG %>%
  select(EVTYPE, Property) %>%
  arrange(desc(Property)) %>%
  head(n=10)

plotPropDMGData
```

- A dataset to analyse crop damage with only the `EVTYPE` and `Crop` columns

```{r tidyDataDMGCrop}
plotCropDMGData <- tidyDataDMG %>%
  select(EVTYPE, Crop) %>%
  arrange(desc(Crop)) %>%
  head(n=10)

plotCropDMGData
```

The following plot shows the impact of weather events on the US economy of property and crop damage side by side:

```{r plotDMGPROPAndCROP, fig.width=9, fig.cap="Fig. 2) Top 10 Weather Events with Economical Impact on (left) Property (right) Crop"}
plotDMGPROP <- 
    ggplot(plotPropDMGData, aes(x = reorder(EVTYPE, -Property), 
                                y = Property/10^9)) +
    geom_bar(stat="identity", fill="blue") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Billions of Dollars", 
        x = "",
      title = "Property")

plotCROPDMG <- 
    ggplot(plotCropDMGData, aes(x = reorder(EVTYPE, -Crop), 
                                y = Crop/10^9)) +
    geom_bar(stat="identity", fill="blue") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Billions of Dollars", 
        x = "",
      title = "Crop")

grid.arrange(plotDMGPROP, plotCROPDMG, ncol = 2,
             main = "Economic impact of Weather Events in the US")

```
  
A new dataset was then created to show the top 20 weather events with the highest impact on both property and crop damage.  
 
```{r tidyDataDMGPropCrop}
plotTotalDMGData <- tidyDataDMG %>%
  arrange(desc(Property), desc(Crop)) %>%
  head(n=20) %>%
  melt(id=c("EVTYPE"), variable.name = "Type")

plotTotalDMGData
```

The plot below shows which event has the most impact on property and crop damages combined.

```{r plotTotalDMG, fig.width=9, fig.cap="Fig. 3) Top 20 Weather Events with Economical Impact (Property and Crop)"}
plotTOTALDMG <- 
    ggplot(plotTotalDMGData, aes(x = reorder(EVTYPE, -value), 
                                 y = value/1000000000, fill = Type)) +
    geom_bar(stat="identity") +
    theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
    labs( 
        y = "Billions of Dollars", 
        x = "",
      title = "Economic impact of Weather Events in the US")

print(plotTOTALDMG)
```

## Conclusions

The conclusions of this analysis are:
* `r tidyDataHarmful[1, "EVTYPE"]` is the Weather Event with the biggest impact on the US Population Health
* `r plotPropDMGData[1, "EVTYPE"]` is the Weather Event with biggest Economical impact on Properties
* `r plotCropDMGData[1, "EVTYPE"]` is the Weather Event with biggest Economical impact on Crops
* `r plotTotalDMGData[1, "EVTYPE"]` and `r plotTotalDMGData[2, "EVTYPE"]` are the Weather Events with biggest Economical impact in the US Economy in general (Property and Crop)

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "NOAA Data Source"
